Water desalination using polyelectrolyte hydrogel: Gibbs ensemble modeling

### 2022-09-16 MDPI/response_letter.tex

\opening{Dear \theeditor,}

In response to your e-mail from 23d August 2022, we hereby submit the revised
version, of our manuscript, entitled "Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modeling''.

We carefully considered all reviewer comments and introduced changes in the manuscript in order to address them.
Below, we provide a point-by-point response to all reviewer comments, including a brief description of the modifications introduced in response to these comments.
Additionally, we provide a pdf version of the manuscript with the relevant changes highlighted in colour.

Main changes in the manuscript include:
\begin{itemize}
    \item Minor changes in title of the article.
    \item The rewritten abstract and introduction.
    \item Added the detailed explanation of model and methods used in the study.
    \item Minor changes in Figures.

\end{itemize}

We also attache the pdf file reflecting the difference between the current version of the manuscript and the submitted one.

We believe that, after considering the modifications in response to reviewer comments, you will find our manuscript suitable for publication in Gels.

@@

\end{letter}

% modify the page to use the space more efficiently
\addtolength{\textwidth}{1.5cm}
\addtolength{\textheight}{2.0cm}

\textbf{Reviewer 1}
\textit{}

\begin{enumerate}
\item \textit{The main findings of your results should be highlighted in abstract}

We have rewritten the abstract.

@@

\item \textit{Please do not separate introduction in three sections combine with problem and technology you used and challenges and your solutions.}

We have rewritten the Introduction section.
We wanted to keep the separation of the Introduction section into the respective subsections because the Introduction is relatively lengthy, so we believe that dividing it into subsections reduces its readability.

@@

\item \textit{In page 4 heading 2. (How to start with and ??)}.\\
Corrected.

\item \textit{All formula and models should be supported with reference for instant equation 3.}
Corrected

\item \textit{In page 6, please remove the footer, it is not a book to describe issues below}.\\
Corrected. Now this sentence appears in text (in brackets).

@@

\item \textit{How do you think control factors in open system such as pressure and temperature}

In our simulation, we control the volume of the simulation box and measure the pressure using the virial theorem. The pressure value which is depicted in the Figures assumed to be a partial pressure of the gel, which is calculated as the difference between the pressure in the gel and the pressure calculated in a separate simulation box containing only ions (with no gel).
Temperature assumed to be constant and equal to 300 K, which is guaranteed by applying a Langevin thermostat.

@@

\end{enumerate}

\textbf{Reviewer 2}
\textit{This manuscript is about creating a new model the thermodynamic equilibrium between coexisting phases of the gel and supernate aqueous solution.}
\begin{enumerate}

\item \textit{Abstract needs to re-write as and please follow the IMRAD instruction.}

We have rewritten the abstract. We believe that the abstract is much more attractive now.
\item \textit{There is introduction in the abstract, however, it needs to write more about the aim of this work, method, result and conclusion.}

Corrected
\item \textit{Avoid to use word 'we' in the abstract as well as manuscript.}

Corrected.
\item \textit{line 247 in what table?}

We meant Table 1. Corrected in text.
\item \textit{Where is the method of this work?}

The method idea is outlined in the paragraph "Physics behind the desalination process" and described in the section "Model and methods". 
%We did not want to bother the readers too much with technical aspects of the smulations, so technical details are in ESI.
\item \textit{It is suggested that to carefully improve this manuscript as well as the writing and grammatical error.}

We improved the writing style of the manuscript and corrected the grammatical errors.

\item \textit{Please explain more detail about the process, how the process run and how the data could be explained and become more interesting for reader.}

Added corresponding paragraph in ``Model and methods `` section

@@

\end{enumerate}

\textbf{Reviewer 3}

\textit{}
\begin{enumerate}
\item \textit{The full stop in the title should be revised.}

Noted. Revised.
\item \textit{The abstract needs to be revised and be more attractive. The research gap, novelty and main findings should be concisely mentioned in the abstract.}

We have rewritten the abstract. We believe that the abstract is much more attractive now.
\item \textit{In this current shape, the manuscript looks like a report. The authors have to revise the flow of the context to attract the readers.}

We have revised and changed the manuscript to be more attractive for the readers. 
\item \textit{The manuscript must be improved by checking once again in English and technical writing. }

We improved the writing style of the manuscript and corrected the grammatical errors.

\item \textit{The whole manuscript must be supported by enough recent references. Revise the whole context.}

We added more recent references.
\item \textit{The beginning of the introduction part must be revised to be more attractive.}

Corrected. We have rewritten the Introduction section.
\item \textit{The authors can follow the following references to generally highlight the global contamination and environmental pollution problems at the beginning of the introduction part \\ \href{https://doi.org/10.1016/j.jwpe.2022.102847}{https://doi.org/10.1016/j.jwpe.2022.102847} ;\\
\href{https://doi.org/10.1016/j.ceramint.2022.05.151}{https://doi.org/10.1016/j.ceramint.2022.05.151} ;\\
\href{https://doi.org/10.3390/catal12050500}{https://doi.org/10.3390/catal12050500} ; \\
\href{https://doi.org/10.1007/s11356-022-21160-7}{https://doi.org/10.1007/s11356-022-21160-7} ;\\
\href{https://doi.org/10.3390/ma15134547}{https://doi.org/10.3390/ma15134547}. \\
Then, highlight the water desalination technologies and Gibbs ensemble\\
(\href{https://doi.org/10.3390/nano10020293}{https://doi.org/10.3390/nano10020293}).}

Corrected. We added the citations, which are relevant to the studied problems, in their appropriate places.
\item \textit{Authors should indicate a clear gap in knowledge which this study seeks to bridge, and potentially contribute to knowledge. }

The purpose of our paper is to show that desalination using polyelectrolyte hydrogels is a cheap and therefore competitive alternative to reverse osmosis.
We highlighted this idea in the Introduction and Conclusion sections.
\item \textit{L 125-126, meaningless sentence. Revise the whole context.}

Corrected.
\item \textit{All sections/subsections titles must be revised to be expressive.}

Corrected. We have rewritten and improved not only the Introduction, but the sections Model and Methods, Results, Conclusions as well.
\item \textit{Statistical physics considerations including setup and calculations should be added.}

We rewritten the section ``Model and methods'' and added there a detailed explanation of our methodology.
\item \textit{The authors should identify the limitations of this study and the recommended future studies in the conclusion part.}

We identified the study limitations and considered the implications and future perspectives of the study. We added the respective paragraphs.
\end{enumerate}

### 2022-09-16 gibbs.tex
finished the revision

\subsection{Langevin Dynamics}
\begin{figure*}[h]
\centering 
\includegraphics[width=0.7\textwidth]{{figures/simgibbs}.pdf}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:diamond}.}
\end{figure*}
We model the gel as a network of 16 linear polymer chains, each consisting of 30 monomer units.
These polymer chains are connected to a diamond-like network by 8 crosslinking units.
This means, there are $\Ngel=16\cdot30 + 8=488$ gel monomers in the simulation box (see \reffig{fig:diamond}).
The network is put in simulation cubic box of the volume $\Vgel$ with periodic boundary conditions, which virtually emulaates an infinite polymer network.

@@

\subsubsection{Monte Carlo sampling in \emph{open system}}
Monte Carlo scheme for sampling the exchange of ions in \emph{open system} is based on the formula for free energy of grand canonical ensemble.
\begin{equation}
    \Omega_{\mathrm{open}}=E-TS+\sum_i\mu_i N_i\label{eq:Omega-GC}
\end{equation}
where $E$ is internal energy, $T$~--- temperature, $S$~--- entropy, $N_i$ is the number of ions of type $i\in\{\na,\ \cl\}$, $\mu_i$~--- corresponding chemical potential.

@@

Using the data obtained from simulations we calculate  
$\cs$ --- the density of ions in  the outside volume;
$\vgel$ --- molar volume of the gel, \ie the volume of the gel per one mol of gel segments, $\vgel = \Vgel / \Ngel$,
$\ncl$ --- the total amount of the ions in both volumes divided by the total volume of both boxes, $\ncl = \Ncl/\Vbox$ and
$\Pgel$ --- partial pressure of the gel, \ie the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a specific molar volume; This pressure is 
We obtain the \emph{partial pressure} as a difference between the presure in gel and in the outer volume $\Pgel=P_\mathrm{gel} - \Pout$.
%$\Pout$, in turn, we obtained from a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$.

@@

\subsection{The efficiency of desalination}
The theoretical minimum specific energy for seawater desalination (
%$\cs=35$ g/l, 
$\cs\simeq0.6$ mol/l as for pure NaCl) is $\sim3.9$ kJ/l (1.1 kWh/m$^3$) for 50\% recovery \cite{Wang_2020}.
This value is calculated as follows 
\begin{equation}
W=2RT\left(\frac{c_{f}}{R_{w}}\ln\frac{c_{b}}{c_{f}}-c_{p}\ln\frac{c_{b}}{c_{p}}\right)
\label{eq:SEC}
\end{equation}
where $R$ is a universal gas constant, $c_{f}$ is the salinity of feed water, $c_p$ is the salinity of product water and $c_b$ is the salinity of the brine, which necessarily appears in any desalination process. 
$R_{w}$ is a recovery ratio, \ie the ratio between volume of produced water and the volume of the feed water. 
50\% recovery ratio means that one part of feed water divides into two equal volume solutions of the product water and of the brine.
Of course, a significant amount of additional energy is required to operate the system\cite{Kim_2019}. 
It has been reported that the specific energy consumption (SEC) of reverse osmosis (RO) process is 2.5~-- 4.0 kWh/m3 (9~-- 14.4 kJ/l), which is significantly higher than its minimum specific energy.
The SEC of a real-scale RO plant is even higher, approximately 3.5~-- 4.5 kWh/$m^{3}$ (12.6 - 16.2 kJ/L), including pre-treatment and post-treatment processes \cite{Kim_2018}. 
%Because of the inherent high energy requirement for RO desalination, seawater is not commonly utilized over traditional surface water.

In order to compare the efficiency of the process presented in \ref{fig: PV and CV} and \ref{fig: NV and CN} with provided values, we collect the corresponding data to the Table~\ref{tab: table}.
The presented desalination process is a cascade of six swellings in open system, at six different (constant) salinities $\cs$, each followed by six compressions in closed system, at six different (constant) $\ncl$.
Each swelling and compression process is presented as a row of Table~\ref{tab: table}, which is colored by matching the lines in the Figures.
The first column of the table contains values $\cs^0$ and $\cs^5$, which stand for the supernate salinity at 0 and 5 bar compression; in open system, supernate salinity does not change, so $\cs^0$ and $\cs^5$ are presented by a single number. 
The second column contains values of $n^0$ and $n^5$, which stand for the number of $\cl$ ions in compression volume $\Vbox$, at 0 and 5 bar pressure (divided by $\Vbox$). 
The number of ion does not change in closed system compression, thus $n^0$ and $n^5$ are the same in the corresponding rows.
The third column shows the change of the gel volume in corresponding process, $\Delta v$. 
The fourth column contains the work needed for the compression in the corresponding process 
per volume of extracted solution. 
This value is an integral of corresponding $\Pgel(\Vgel)$ dependence \citep{Atkins} 
\begin{equation}
    W = \frac{\int_{v^0}^{v^5} \Pgel d\Vgel}{\Delta v}
\end{equation}
Here, in the table, we present the absolute values of the work, 
whereas one should keep in mind that the compression implies the work 
is done by external force and the swelling implies that the work is done by gel.

@@

To give a clue of comparison with ideal desalination process efficiency, the fifth column provides the values of specific energy consumption, $W^\text{id}$, which are calculated by means of \refeq{eq:SEC}, using the following procedure:
Let§s assume the concentration of the feed solution to be $\cs^f=44.91$ mmol/l and the concentrations of product and brine solutions to be $\cs^p=31.93$ mmol/l and $\cs^b = 63.93$ mmol/l, respectively.
\begin{enumerate}
    \item The gel, equilibrated with the feed solution, is compressed in the \emph{closed system}.
    The gel volume decreases by $\Delta v^p = 3.69$ liters and the salinity of supernate decreases from $\cs^f$ to $\cs^p$.
    The volume of the product solution is $\Delta v^p$
    \item The squeezed gel is put back into the feed solution and is quilibrated there under pressure, so it does not swell. 
    \item After the equilibration, the gel swells in the \emph{closed system} and the salinity of the external solution increases to the value $\cs^b$.
    \item Finally, the gel is taken out and compressed by 5 bar in the \emph{open system} in equilibrium with the brine bath. 
    The change of the gel volume in this process is $\Delta v^b = 3.26$ l/mol, which equals to the volume of the produced brine. 
\end{enumerate}
Thus the recovery ratio $R_w = \Delta v^p / (\Delta v^p + \Delta v^b) \simeq $ 0.53 and the theoretical minimum specific energy of the desalination process with corresponding $\cs^f$, $\cs^p$ and $\cs^b$ is $W^{id} =$ 38.2 J/l (\refeq{eq:SEC}).

The estimated $W^{id}$ values are provided in the fifth column of the table, for five triplets of $\cs^f$, $\cs^p$,$\cs^b$ values .
$W_{\mathrm{sim}}$ is the energy calculated by integration of $\Pgel(\Vgel)$ dependencies as the sum of corresponding compressions in \emph{closed} and \emph{open} systems, $W_{\mathrm{id}}$ is the estimation based on the equation \refeq{eq:SEC} and $R_w$ is the corresponding recovery ratio.
\todo{how the integration was performed}
Note that calculating the $W_{\mathrm{sim}}$ we accounted for only the work done on the gel during the compression, whereas the work done by the gel itself during swelling was not taken into account.
The ratio between $W_{\mathrm{sim}}$ and $W_{\mathrm{id}}$ ranges from 3.26 to 5.43, which is comparable to that of RO.

>>>>

\subsection{Molecular Dynamics}
\begin{figure*}[h]
\centering 
\includegraphics[width=0.7\textwidth]{{figures/simgibbs}.pdf}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:diamond}.}
\end{figure*}
We model the gel as a network of 16 linear polymer chains, each consisting of 30 monomer units.
These polymer chains are connected to a diamond-like network by 8 crosslinking units.
This means, there are $\Ngel=16\cdot30 + 8=488$ gel monomers in the simulation box (see \reffig{fig:diamond}).
The network is put in simulation cubic box of the volume $\Vgel$ with periodic boundary conditions, which virtually emulaates an infinite polymer network.

@@

\subsection{Monte Carlo sampling in open system}
Monte Carlo scheme for sampling the exchange of ions in \emph{open system} is based on the formula for free energy of grand canonical ensemble, $\Omega$
\begin{equation}
    \Omega_{\mathrm{open}}=E-TS+\sum_i\mu_i N_i\label{eq:Omega-GC}
\end{equation}
where $E$ is internal energy, $T$~--- temperature, $S$~--- entropy, $N_i$ is the number of ions of type $i\in\{\na,\ \cl\}$, $\mu_i$~--- corresponding chemical potential.

@@

Using the data obtained from simulations we calculate  
$\cs$ --- the density of ions in  the outside volume;
$\vgel$ --- molar volume of the gel, \ie the volume of the gel per one mol of gel segments, $\vgel = \Vgel / \Ngel$,
$\ncl$ --- the total amount of the ions in both volumes divided by the total volume of both boxes, $\ncl = \Ncl/\Vbox$ and
$\Pgel$ --- \emph{partial pressure} of the gel, \ie the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a specific molar volume.
We obtain the gel partial pressure as a difference between the presure in gel and in the outer volume $\Pgel=P_\mathrm{gel} - \Pout$.
%$\Pout$, in turn, we obtained from a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$.

@@

\subsection{The efficiency of desalination}
The theoretical minimum specific energy for seawater desalination ($\cs\simeq0.6$ mol/l as for pure NaCl) is $\sim3.9$ kJ/l (1.1 kWh/m$^3$) for 50\% recovery \cite{Wang_2020}.
This value is calculated as follows
\begin{equation}
W_{\mathrm{id}}=2RT\left(\frac{c_{f}}{R_{w}}\ln\frac{c_{b}}{c_{f}}-c_{p}\ln\frac{c_{b}}{c_{p}}\right)
\label{eq:SEC}
\end{equation}
where $R$ is a universal gas constant, $c_{f}$ is the salinity of feed water, $c_p$ is the salinity of product water and $c_b$ is the salinity of the brine, which necessarily appears in any desalination process.
$R_{w}$ is a recovery ratio, \ie the ratio between the volume of produced water and the volume of the feed water.
50\% recovery ratio means that one part of feed water divides into two equal volume solutions of the product water and of the brine.
Of course, a significant amount of additional energy is required to operate the system\cite{Kim_2019}.
It has been reported that the specific energy consumption (SEC) of the reverse osmosis (RO) process is 2.5~-- 4.0 kWh/m3 (9.0~-- 14.4 kJ/l), which is significantly higher than its minimum specific energy.
The SEC of a real-scale RO plant is even higher, approximately 3.5~-- 4.5 kWh/$m^{3}$ (12.6 - 16.2 kJ/l), including pre-treatment and post-treatment processes \cite{Kim_2018}.
%Because of the inherent high energy requirement for RO desalination, seawater is not commonly utilized over traditional surface water.

To compare the efficiency of the desalination process presented in \reffig{fig: PV and CV} and \reffig{fig: NV and CN} with provided values, we collect the corresponding data to Table~\ref{tab: table}.
The presented desalination process is a cascade of six swellings in an open system, at six different (constant) salinities $\cs$, each followed by six compressions in a closed system, at six different (constant) $\ncl$.
Each swelling and compression process is presented as a row of Table~\ref{tab: table}, which is colored by matching the lines in the Figures.
The first column of the table contains values $\cs^0$ and $\cs^5$, which stand for the supernate salinity at 0 and 5 bar compression; in an open system, supernate salinity does not change, so $\cs^0$ and $\cs^5$ are presented by a single number.
The second column contains values of $n^0$ and $n^5$, which stand for the number of $\cl$ ions in compression volume $\Vbox$, at 0 and 5 bar pressure (divided by $\Vbox$).
The number of ions does not change in closed system compression, thus $n^0$ and $n^5$ are the same in the corresponding rows.
The third column shows the change of the gel volume in the corresponding process, $\Delta v$.
The fourth column contains the work needed for the compression in the corresponding process per volume of extracted solution.
This value is obtained as numerical integration of corresponding $\Pgel(\Vgel)$ dependence \citep{Atkins}
\begin{equation}
W = \frac{\int_{v^0}^{v^5} \Pgel d\Vgel}{\Delta v}
\end{equation}
In this column, we presented the absolute values of the work,
whereas one should keep in mind that compression implies, that the work is done by external force and swelling implies, that the work is done by gel.

@@

To give a clue of comparison with ideal desalination process efficiency, the fifth column provides the values of ideal specific energy consumption, $W^\text{id}$, which are calculated employing \refeq{eq:SEC}, for concentrations of feed, product and brine solutions, $\cs^f$,  $\cs^p$, $\cs^b$, as indicated by curly brackets.
For example, having $\cs^f=44.91$, $\cs^p=31.93$ and $\cs^b = 63.93$ mmol/l, (5th, 7th and 3rd rows of the table) one can imagine the following desalination process
\begin{enumerate}
\item First, the gel equilibrated with the feed solution, is compressed in the \emph{closed system}.
The gel volume decreases by $\Delta v^p = 3.69$ liters and the salinity of supernate decreases from $\cs^f$ to $\cs^p$.
The volume of the product solution is $\Delta v^p$
\item The squeezed gel is put back into the feed solution and is equilibrated there under pressure, so it does not swell.
\item After the equilibration, the gel swells in the \emph{closed system}, so the salinity of the external solution increases to the value $\cs^b$.
\item Finally, the gel is taken out and compressed by pressure 5bar in the \emph{open system} in equilibrium with the brine bath.
The change of the gel volume in this process is $\Delta v^b = 3.26$ l/mol, which equals the volume of the produced brine.
\end{enumerate}
Thus the recovery ratio $R_w = \Delta v^p / (\Delta v^p + \Delta v^b) \simeq $ 0.53 and the theoretical minimum specific energy of the desalination process with corresponding $\cs^f$, $\cs^p$ and $\cs^b$ is $W^{id} =$ 38.2 J/l (\refeq{eq:SEC}).

The estimated $W^{id}$ values are provided in the fifth column of the table, for five triplets of $\cs^f$, $\cs^p$,$\cs^b$ values.
In the same column we provide $W_{\mathrm{sim}}$ --- the specific energy consumptions calculated by a numerical integration of $\Pgel(\Vgel)$ dependencies. 
The provided values  $W_{\mathrm{sim}}$ are the sum of energies needed for corresponding compression processes, \ie in \emph{closed} and \emph{open} systems.
The values $R_w$, which are also provided in 5th colum are the corresponding recovery ratios.

The ratio between $W_{\mathrm{sim}}$ and $W_{\mathrm{id}}$ ranges from 3.26 to 5.43, which is comparable to that of RO.
Note, that calculating the $W_{\mathrm{sim}}$ we accounted for only the work done on the gel during the compression, whereas the work done by the gel itself during swelling was not taken into account. 
The process, which accounts for the energy recovery was described in our previous studies \cite{Prokacheva2021, Rud2018}
